Historic collections as a tool for assessing the global pollination crisis

### 2018-06-15 revision/ms.Rmd
Revised ms.

---
title: Historic collections as a tool for assessing the global pollination crisis
author: "Bartomeus, I.^1^, Stavert, J.R.^2^, Ward, D.^2,3^, and Aguado, O.^4^"
csl: data/philb_style.xml
output:
  html_document: default
  word_document: default
editor_options:
  chunk_output_type: console
bibliography: data/references2.xml
---

@@

One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and lizards [@Ollerton2017]. Additionally, many pollinators are highly mobile, short-lived and small in size, which makes monitoring their populations difficult. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Winfree2011]. However, bees are diverse, with more than 20,000 species currently described worldwide, and often require expert taxonomists for identification. Furthermore, the uneven global distribution of research studies has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].

@@

```{r map, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(rworldmap)
map <- read.csv("data/bee_richness.csv")
size <- read.csv("data/country_area.csv")
colnames(size) <- c("country", "size")
map0 <- merge(map, size, all.x = TRUE)
#remove some countries with doubtful sampling
#map0 <- subset(map0, !country %in% c("Kuwait", "Guinea-Bissau", "Chad", "Mali", "Mauritania"))
#map0[which(is.na(map0$size)),] #some can be rescued?
#scatter.smooth(log(map0$species) ~ log(map0$size+1))
m <- lm(log(map0$species) ~ log(map0$size+1), na.action = na.exclude) 
map0$res <- residuals(m)
#subset(map0, res < -2)
map2 <- joinCountryData2Map(map0, joinCode="NAME", nameJoinColumn = "country",
                            nameCountryColumn = "distribution")
#need to use ISO's to make sure we don't carry over NA's due to country misspellings
#have plotted countries with missing values as grey (shows we have richness for pretty much everywhere)
par(mfrow = c(2,2))
map3 <- mapCountryData(map2, nameColumnToPlot = "res", missingCountryCol = "white",
               mapTitle = "Bee species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE)
#add a modified legend using the same initial parameters as mapCountryData
do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5))

@@

#read in data
declines <- read.csv(file = "data/richness_declines.V2.csv", header = TRUE, sep = ",", stringsAsFactors=FALSE,na.strings=c("","NA"))
declines$percent_richness_change <- as.numeric(declines$percent_richness_change)
declines$proportion_richness_change <- as.numeric(declines$proportion_richness_change)
#nothing outside EU- USA. Spain and NZ can be added.
#head(declines)

@@

#average richness change
map.bee.ave <- mapCountryData(map.bee, nameColumnToPlot = "average", colourPalette="topo",
                                  mapTitle = "Average bee richness change", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.bee.ave, legendLabels="all", legendWidth=0.5))

@@

#average richness change
map.syrphid.ave <- mapCountryData(map.syrphid, nameColumnToPlot = "average", colourPalette="topo",
                       mapTitle = "Average syrphid richness change", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.syrphid.ave, legendLabels="all", legendWidth=0.5))

@@

#average richness change
map.butterfly.ave <- mapCountryData(map.butterfly, nameColumnToPlot = "average", colourPalette="topo",
                                    mapTitle = "Average butterfly richness change", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.butterfly.ave, legendLabels="all", legendWidth=0.5))

@@

```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
#lhf
#head(lhf)
#str(lhf)
lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
#write.csv(lhf, "data/SpMat2.csv")
#plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
#abline(h = 150000, v = 150000)
#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.

@@

```{r map rgbif, echo=FALSE, message=FALSE, warning=FALSE}
countries <- lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),"country"]
load("data/gbif.RData")
#3) split by year
library(vegan)
post1980 <- c()
pre1980 <- c()
n <- c()
for(i in 1:length(countries)){
    dat <- occ_list[[i]]
    #nrow(dat) 
    dat2 <- dat[-which(is.na(dat$name)),]
    dat3 <- dat2[which(is.na(as.numeric(dat2$name))),]
    dat4 <- subset(dat3, name != "Apis mellifera")
    #nrow(dat1) 
    dat1 <- dat4
    #unique(dat1$name)
    dat1$period <- ifelse(dat1$year < 1980, "pre1980", "post1980")
    #table(dat1$period) # :(
#4) rar richness
    comm <- dcast(dat1, period ~ name, fun.aggregate = length, value.var = "name")
    rownames(comm) <- comm$period
    comm <- comm[,-1]
    temp <- rarefy(comm, sample = round(min(table(dat1$period))*0.75))
    #str(temp)
    post1980[i] <- temp[1]
    pre1980[i] <- temp[2]
    n[i] <- round(min(table(dat1$period))*0.75)
}
#5) t test!
data <- data.frame(countries = as.character(countries), pre1980 = pre1980, post1980 = post1980, n = n)
#head(data)
#maybe delete n < 500?
data2 <- subset(data, n > 500) #28 countries with enought data.

@@

#average richness change
#colorss <- c("firebrick", "red", "orangered", "orange", "yellow", "wheat", "green", "blue", "darkblue")
colorss <- (heat.colors(10))
map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette= colorss,
                       mapTitle = "Average bee richness change (%)", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.gbif.plot, legendLabels="all", legendWidth=0.5))

@@

```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
dat <- read.csv(file = "data/spain_clean_used.csv")
#Repeat richness and genus
rich_tot <- dcast(dat, when ~ ., value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#rich_tot #wow.
#extract genus
pos <- regexpr(dat$matched_name2, pattern = " ", fixed = TRUE)
dat$genus <- substr(dat$matched_name2, 
                    start = 1,
                    stop = pos-1)
genus_tot <- dcast(dat, when ~ ., value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
genus_tot2 <- dcast(dat,  genus ~ when, value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot #wow, lot's of genus lost!
#genus_tot2
#subset(genus_tot2, now == 0 & asensio == 1) #20 lost!
#subset(genus_tot2, now == 1 & asensio == 1) #27 stable
#subset(genus_tot2, now == 1 & asensio == 0) #2 added!
genus_tot3 <- dcast(dat,  genus ~ when, value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot3
#colSums(genus_tot3[,2:3])
#write.csv(genus_tot3, "data/SupMat4.csv")

@@

d <- read.csv("data/NZBees_allrecords.csv", header = TRUE, sep = ",", na.strings=c("","NA"))

@@

    #load data
    d <- read.csv(file = "data/NZ_fly_pollinators.csv", header = TRUE, sep = ",", na.strings=c("","NA"))

>>>>

---
title: "Historic collections as a tool for assessing the global pollination crisis"
author: "Bartomeus, I.^1^, Stavert, J.R.^2^, Ward, D.^2,3^, and Aguado, O.^4^"
csl: ../data/philb_style.xml
output:
  word_document: default
  html_document: default
editor_options:
  chunk_output_type: console
bibliography: ../data/references2.xml
---

@@

One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and lizards [@Ollerton2017]. Additionally, many pollinators are highly mobile, short-lived and small in size, which makes monitoring their populations difficult. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Winfree2011]. However, bees are diverse, with more than 20,000 species currently described worldwide [Asher], and often require expert taxonomists for identification. Furthermore, the uneven global distribution of research studies has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].

@@

```{r map, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(rworldmap)
map <- read.csv("../data/bee_richness.csv")
size <- read.csv("../data/country_area.csv")
colnames(size) <- c("country", "size")
map0 <- merge(map, size, all.x = TRUE)
#remove some countries with doubtful sampling
#map0 <- subset(map0, !country %in% c("Kuwait", "Guinea-Bissau", "Chad", "Mali", "Mauritania"))
#map0[which(is.na(map0$size)),] #some can be rescued?
#scatter.smooth(log(map0$species) ~ log(map0$size+1))
m <- lm(log(map0$species) ~ log(map0$size+1), na.action = na.exclude) 
map0$res <- residuals(m)
#subset(map0, res < -2)
map2 <- joinCountryData2Map(map0, joinCode="NAME", nameJoinColumn = "country",
                            nameCountryColumn = "distribution")
#need to use ISO's to make sure we don't carry over NA's due to country misspellings
#have plotted countries with missing values as grey (shows we have richness for pretty much everywhere)
par(mfrow = c(2,2))
map3 <- mapCountryData(map2, nameColumnToPlot = "res", missingCountryCol = "white",
               mapTitle = "Bee species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE)
#add a modified legend using the same initial parameters as mapCountryData
do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5))

@@

#read in data
declines <- read.csv(file = "../data/richness_declines.V2.csv", header = TRUE, sep = ",", stringsAsFactors=FALSE,na.strings=c("","NA"))
declines$percent_richness_change <- as.numeric(declines$percent_richness_change)
declines$proportion_richness_change <- as.numeric(declines$proportion_richness_change)
#nothing outside EU- USA. Spain and NZ can be added.
#head(declines)

@@

#average richness change
colorss <- c(heat.colors(9))
map.bee.ave <- mapCountryData(map.bee, nameColumnToPlot = "average", colourPalette= colorss,
                                  mapTitle = "Average bee richness change", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.bee.ave, legendLabels="all", legendWidth=0.5))

@@

#average richness change
colorss <- c(heat.colors(5), c("#5e977c", "#3d88a5", "#1c79cf"))
map.syrphid.ave <- mapCountryData(map.syrphid, nameColumnToPlot = "average", colourPalette= colorss,
                       mapTitle = "Average syrphid richness change", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.syrphid.ave, legendLabels="all", legendWidth=0.5))

@@

#average richness change
colorss <- c(heat.colors(6))
map.butterfly.ave <- mapCountryData(map.butterfly, nameColumnToPlot = "average", colourPalette=colorss,
                                    mapTitle = "Average butterfly richness change", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.butterfly.ave, legendLabels="all", legendWidth=0.5))

@@

```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
lhf <- read.csv("../data/rgbif_output.csv", row.names = 1 )
#lhf
#head(lhf)
#str(lhf)
lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
#write.csv(lhf, "data/SpMat2.csv")
#plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
#abline(h = 150000, v = 150000)
#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.

@@

```{r map rgbif, echo=FALSE, message=FALSE, warning=FALSE}
countries <- lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),"country"]
load("../data/gbif.RData")
#3) split by year
library(vegan)
post1980 <- c()
pre1980 <- c()
n <- c()
for(i in 1:length(countries)){
    dat <- occ_list[[i]]
    #nrow(dat) 
    dat2 <- dat[-which(is.na(dat$name)),]
    dat3 <- dat2[which(is.na(as.numeric(dat2$name))),]
    dat4 <- subset(dat3, name != "Apis mellifera")
    #nrow(dat1) 
    dat1 <- dat4
    #unique(dat1$name)
    dat1$period <- ifelse(dat1$year < 1980, "pre1980", "post1980")
    #table(dat1$period) # :(
#4) rar richness
    comm <- dcast(dat1, period ~ name, fun.aggregate = length, value.var = "name")
    rownames(comm) <- comm$period
    comm <- comm[,-1]
    temp <- rarefy(comm, sample = round(min(table(dat1$period))*0.75))
    #str(temp)
    post1980[i] <- temp[1]
    pre1980[i] <- temp[2]
    n[i] <- round(min(table(dat1$period))*0.75)
}
#5) t test!
data <- data.frame(countries = as.character(countries), pre1980 = pre1980, post1980 = post1980, n = n)
#head(data)
#maybe delete n < 500?
data2 <- subset(data, n > 500) #28 countries with enought data.

@@

#average richness change
#colorss <- c("firebrick", "red", "orangered", "orange", "yellow", "wheat", "green", "blue", "darkblue")
colorss <- c(heat.colors(5), c("#5e977c", "#3d88a5", "#1c79cf"))
map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette= colorss,
                       mapTitle = "Average bee richness change (%)", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.gbif.plot, legendLabels="all", legendWidth=0.5))

@@

```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
dat <- read.csv(file = "../data/spain_clean_used.csv")
#Repeat richness and genus
rich_tot <- dcast(dat, when ~ ., value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#rich_tot #wow.
#extract genus
pos <- regexpr(dat$matched_name2, pattern = " ", fixed = TRUE)
dat$genus <- substr(dat$matched_name2, 
                    start = 1,
                    stop = pos-1)
genus_tot <- dcast(dat, when ~ ., value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
genus_tot2 <- dcast(dat,  genus ~ when, value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot #wow, lot's of genus lost!
#genus_tot2
#subset(genus_tot2, now == 0 & asensio == 1) #20 lost!
#subset(genus_tot2, now == 1 & asensio == 1) #27 stable
#subset(genus_tot2, now == 1 & asensio == 0) #2 added!
genus_tot3 <- dcast(dat,  genus ~ when, value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot3
#colSums(genus_tot3[,2:3])
#write.csv(genus_tot3, "data/SupMat4.csv")

@@

d <- read.csv("../data/NZBees_allrecords.csv", header = TRUE, sep = ",", na.strings=c("","NA"))

@@

    #load data
    d <- read.csv(file = "../data/NZ_fly_pollinators.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
